Unrestricted Termination and Non-Termination Arguments for Bit-Vector Programs

### 2015-04-12 slides/esop-termination.tex
Non termination

\begin{frame}{Recurrence Sets}
 To prove that a loop terminates, we tried to find a ranking function.

@@

 In order to prove that a loop does \emph{not} terminate, we will exhibit a \emph{recurrence set}.

 \vspace{1em}
\end{frame}

>>>>

\begin{frame}{Recurrence Sets}
 To prove that a loop $(P, G, B)$ terminates, we tried to find a ranking function.

@@

 In order to prove that a loop does \emph{not} terminate, we will exhibit a \emph{recurrence set} --
 this is a non-empty set of states $S$ that, if entered, will never be left:
 \begin{align*}
  \exists x_0 . & S(x_0) \\
  \forall x . S(x) \Rightarrow & G(x) \\
  \forall x, x' . S(x) \wedge B(x, x') \Rightarrow & S(x')
 \end{align*}

\begin{frame}[fragile]{Recurrence Set Example}
\begin{exampleblock}{Example}
\begin{center}
\begin{columns}[c]
 \column{.3\textwidth}
  \begin{tabular}{c}
   \begin{lstlisting}[language=c,basicstyle=\normalsize]
while (x > 0) {
  x = x / 2;
  x = x + 1;
}
   \end{lstlisting}
  \end{tabular}
 \column{.3\textwidth}
 \begin{gather*}
  x_0 = 2 \\
  S = \{ 2 \}
 \end{gather*}
\end{columns}
\end{center}
\end{exampleblock}
\end{frame}

\begin{frame}[fragile]{Nondeterminism and Non-Termination}
 If the loop body contains some nondeterminism, we only require that for each $\red{S}$-state, there is
 \emph{some} successor that is an $\red{S}$-state, rather than requiring that \emph{every} successor is an
 $\red{S}$-state:
  \begin{align*}
  \exists \green{x_0} . & \red{S}(\green{x_0}) \\
  \forall x . \red{S}(x) \Rightarrow & G(x) \\
  \forall x \exists x' . \red{S}(x) \wedge B(x, x') \Rightarrow & \red{S}(x')
 \end{align*}

Our syntax does not allow us to existentially quantify over ground terms inside the universal.  To
get around this, we Skolemise the formula, introducing a successor function $\blue{N}$ which picks
out the next state for each $\red{S}$-state.
\end{frame}

\begin{frame}{Second-Order Non-Termination}

\begin{block}{Skolemised Non-Termination Formula [SNT]}
 \[
 \exists \green{x_0}, \red{S}, \blue{N} . \forall x . \left ( \begin{array}{rl}
   \red{S}(\green{x_0}) \, \wedge & \\
   \red{S}(x) \Rightarrow & G(x) \, \wedge \\
   \red{S}(x) \Rightarrow & B(x, \blue{N}(x)) \wedge \red{S}(\blue{N}(x)) 
   \end{array} \right )
 \]
 \end{block}

This formula is satisfiable iff there exists a recurrence set $\blue{S}$ containing
an initial state $\red{x_0}$, with successor function $\green{N}$, i.e. iff the loop
does not terminate for some input $\red{x_0}$.
\end{frame}
